---
title: "Predict Future Elk Harvest"
author: "Pierre Sarnow"
output:
  html_notebook:
    toc: yes
    toc_float: false
    toc_depth: 6
    theme: yeti
    hightlight: default
    code_folding: none
---
***
## Description
Use historical harvests, number of hunters, and weather data to predict the harvest for the upcoming hunting seasons.

*__NOTICE__ that I am only looking at the general rifle hunting seasons on public land. There are also 
hunters for Archery, Muzzleloader, Private Land, Ranching for Wildlife, etc.*

***
## Setup
Load required libraries for wrangling data, charting, and mapping
```{r}
library(plyr,quietly = T) # data wrangling
library(dplyr,quietly = T) # data wrangling
library(ggplot2, quietly = T) # charting
library(ggthemes,quietly = T) # so I can add the highcharts theme and palette
library(scales,quietly = T) # to load the percent function when labeling plots
library(caret,quietly = T) # classification and regression training
library(foreach,quietly = T) # parallel processing to speed up the model training
library(doMC,quietly = T) # parallel processing to speed up the model training
library(lubridate,quietly = T) # for timing models
```

Set our preferred charting theme
```{r}
theme_set(theme_minimal()+theme_hc()+theme(legend.key.width = unit(1.5, "cm")))
``` 

Run script to get hunter data
```{r}
source('~/_code/colorado-dow/datasets/Colorado Elk Harvest Data.R', echo=F)
```

Table of the harvest data
```{r}
head(COElkRifleAll)
```


Load the weather data
```{r}
load("weatherdata5.RData")
head(weatherdata5)
```

source geodata
```{r}
source('~/_code/colorado-dow/datasets/Colorado GMUnit and Road data.R', echo=F)
```

Take a peak at the boundary data
```{r}
head(Unitboundaries2)
```

Set to predictive analytics directory
```{r}
setwd("~/_code/colorado-dow/phase III - predictive analytics")
```
***
### Organize data
I could also use the harvest and population estimates to create another elk population number, might be useful
in predicting. 
Because there are some elk not accounted for in DecemberPopulation = JanuaryPopulation - FallHarvest

Group weather data by year
```{r}
UnitWeather <- summarise(group_by(weatherdata5,Year,Unit),
                         daily.temperatureHigh = max(daily.temperatureHigh,na.rm = T),
                         daily.temperatureLow = min(daily.temperatureLow,na.rm = T),
                         daily.temperatureMean = mean(daily.temperatureMean,na.rm = T),
                         daily.precipAccumulation = mean(daily.precipAccumulation,na.rm = T),
                         daily.precipType = mean(daily.precipType,na.rm = T),
                         daily.windSpeed = mean(daily.windSpeed,na.rm = T),
                         daily.FullmoonPhase = mean(daily.FullmoonPhase,na.rm = T))
```

Appropriate field classes for model training
```{r}
UnitWeather$Year <- as.numeric(UnitWeather$Year)
```

Group Hunter data by Year and Unit
```{r}
COElkHarvest <- summarise(group_by(COElkRifleAll,Year,Unit),
                          Harvest = sum(c(Harvest.Antlered,Harvest.Antlerless),na.rm = T),
                          Hunters = sum(c(Hunters.Antlered,Hunters.Antlerless,Hunters.Either),na.rm = T))

COElkHarvest$Year <- as.numeric(COElkHarvest$Year)
```

Join Harvest, Hunter, and Weather data
```{r}
COElkHarvest <- full_join(COElkHarvest, UnitWeather, by = c("Year","Unit"))
```

Remove rows with missing data
```{r}
COElkHarvest <- filter(COElkHarvest, !is.na(Harvest) & !is.na(daily.temperatureMean) & Year != 2018)
```

Split into train and test sets. Will use 75% of the data to train on. 

```{r}
COElkHarvest <- mutate(group_by(COElkHarvest,Unit),
                       numentries = n())
COElkHarvest <- filter(COElkHarvest, numentries >= 3)
COElkHarvest$UnitYear <- paste(COElkHarvest$Unit, COElkHarvest$Year)

traindata <- COElkHarvest %>% group_by(Unit) %>% sample_frac(size = .75, replace = F)
testdata <- COElkHarvest[!COElkHarvest$UnitYear %in% traindata$UnitYear,]

COElkHarvest <- select(COElkHarvest, -UnitYear, -numentries)

traindata <- select(traindata, -UnitYear, -numentries)
testdata <- select(testdata, -UnitYear, -numentries)
```

Save off for importing into AzureML
```{r}
write.csv(COElkHarvest,file = "~/_code/colorado-dow/datasets/COElkHarvest.csv",row.names = F)
```

***
## Model Building

### Model Training Methods
Loop through possible methods, utilizing the quicker 'adaptive_cv' parameter search from caret.
Consider scripting this into AzureML to make it run much faster, though there is more setup and errors to 
control for

```{r}
quickmethods <- c("lm",'svmLinear',"svmRadial","knn","cubist","kknn")

step1_all <- NULL
for (imethod in quickmethods) {
  step1 <- NULL
  start <- now()
  
  if (imethod == "lm" | imethod == "svmLinear") {
    controlmethod <- "repeatedcv"
  } else {controlmethod <- "adaptive_cv"}
  
  fitControl <- trainControl(
    method = controlmethod,
    # search = 'random',
    number = 4,
    repeats = 4,
    allowParallel = TRUE,
    summaryFunction = defaultSummary)
  
  registerDoSEQ()
  registerDoMC(cores = 6)
  
  HarvestModel_1 = train(Harvest ~ ., data = traindata,
                         method = imethod,
                         #preProc = c("center","scale"), 
                         tuneLength = 15,
                         trControl = fitControl)
  
  HarvestModel_1
  
  # measure performance
  predictdata <- predict(HarvestModel_1, testdata)
  
  step1$method <- imethod
  step1$RMSE <- postResample(pred = predictdata, obs = testdata$Harvest)[1]
  step1$duration <- now() - start
  step1 <- as.data.frame(step1)
  step1_all <- rbind(step1_all,step1)
}
```
View Results
```{r}
step1_all
```
Run the best model with all of the data
```{r}
fitControl <- trainControl(
    method = "repeatedcv",
    # search = 'random',
    number = 4,
    repeats = 4,
    allowParallel = TRUE,
    summaryFunction = defaultSummary)
  
registerDoSEQ()
registerDoMC(cores = 6)
  
HarvestModel_1 = train(Harvest ~ ., data = COElkHarvest,
                         method = 'lm',
                         #preProc = c("center","scale"), 
                         tuneLength = 15,
                         trControl = fitControl)
  
HarvestModel_1
```

## Predict Hunter for next year, 2018
```{r}
# Get 2018 Hunters from the Hunters Predicted model
load("~/_code/colorado-dow/datasets/COElkHunters2018_Season.RData")

COElkHunters2018Predicted <- summarise(group_by(COElkHunters2018_Season, Year, Unit),
                                       Hunters = sum(Hunters,na.rm = T))

# ensure we include all of the units
COElkHarvest2018 <- as.data.frame(unique(COElkHarvest$Unit))
colnames(COElkHarvest2018) <- "Unit"
COElkHarvest2018$Year <- 2018

# Weather data for 2018
UnitWeather2018 <- filter(UnitWeather,Year==2018)

# A left join will autofill missing draw data with NAs, but will retain the full list of Units
COElkHarvest2018 <- left_join(COElkHarvest2018,UnitWeather2018)

# Join in forecasted Hunters
COElkHarvest2018 <- left_join(COElkHarvest2018,COElkHunters2018Predicted)
# Replace the hunter data with missing entries
COElkHarvest2018$Hunters[is.na(COElkHarvest2018$Hunters)] <- 0

COElkHarvest2018 <- COElkHarvest2018[, colnames(COElkHarvest2018) %in% c("Unit",HarvestModel_1$coefnames)]

COElkHarvest2018$Harvest <- predict(HarvestModel_1, COElkHarvest2018)
COElkHarvest2018$Harvest[COElkHarvest2018$Harvest<0] <- 0

# Combine with historic data
COElkHarvestAll <- rbind.fill(COElkHarvest,COElkHarvest2018)
```
Save off so we don't have to recreate the model everytime we want the results
```{r}
save(COElkHarvest2018,file="~/_code/colorado-dow/datasets/COElkHarvest2018.RData")
```

***
## View results
```{r fig.width=10}
ggplot(COElkHarvestAll, aes(Year,Harvest)) +
  geom_bar(position="dodge",stat="identity") +
  # coord_cartesian(ylim = c(130000,155000)) +
  scale_fill_hc() +
  labs(title="Statewide Elk Harvest", caption="source: cpw.state.co.us")
```

```{r fig.width=10}
Harvest_77 <- filter(COElkHarvestAll, Unit == "77")
ggplot(Harvest_77, aes(Year,Harvest)) +
  geom_bar(position="dodge",stat="identity") +
  # coord_cartesian(ylim = c(130000,155000)) +
  scale_fill_hc() +
  labs(title="Unit 77 Elk Harvest", caption="source: cpw.state.co.us")
```


