---
title: "Predict Number of Future Elk Hunters"
author: "Pierre Sarnow"
output:
  html_document:
    toc: yes
    toc_float: false
    theme: yeti
    hightlight: default
    fig_width: 10
---


***
## Description
Use historical draw results, and number of hunters to train a model we can use to 
predict the number of hunters in future years.

TODO - Include other potential inputs that could impact how many hunters get a license
and show up. Those could include economic indicators, and costs associated with hunting
(transportation, lodging).

*__NOTICE__ that I am only looking at the general rifle hunting seasons on public land. There are also 
hunters for Archery, Muzzleloader, Private Land, Ranching for Wildlife, etc.*

***
## Setup
Load required libraries for wrangling data, charting, and mapping
```{r}
library(plyr,quietly = T) # data wrangling
library(dplyr,quietly = T) # data wrangling
library(ggplot2, quietly = T) # charting
library(ggthemes,quietly = T) # so I can add the highcharts theme and palette
library(scales,quietly = T) # to load the percent function when labeling plots
library(caret,quietly = T) # classification and regression training
library(foreach,quietly = T) # parallel processing to speed up the model training
library(doMC,quietly = T) # parallel processing to speed up the model training
library(lubridate,quietly = T) # for timing models
```

Set our preferred charting theme
```{r}
theme_set(theme_minimal()+theme_hc()+theme(legend.key.width = unit(1.5, "cm")))
```

Run script to get hunter data
```{r}
source('~/_code/colorado-dow/datasets/Colorado Elk Harvest Data.R', echo=F)
```

Table of the harvest data
COElkRifleAll

Run script to get draw data
```{r}
source('~/_code/colorado-dow/datasets/Elk Drawing Summaries.R', echo=F)
```

Table of the data
```{r}
head(COElkDrawAll)
```

source geodata
```{r}
source('~/_code/colorado-dow/datasets/Colorado GMUnit and Road data.R', echo=F)
```

Take a peak at the boundary data
```{r}
head(Unitboundaries2)
```

Set to predictive analytics directory
```{r}
setwd("~/_code/colorado-dow/phase III - predictive analytics")
```

***
## Organize data
Will start by grouping all of the seasons together, and modeling the number of hunters per
Year and Unit
Group Draw results data by Year and Unit
```{r}
COElkDraw <- summarise(group_by(COElkDrawAll,Year,Unit),
                       Quota = sum(Orig_Quota,na.rm = T),
                       Drawn = sum(Chcs_Drawn,na.rm = T))
```

Appropriate field classes for model training
```{r}
COElkDraw$Year <- as.numeric(COElkDraw$Year)
```

Group Hunter data by Year and Unit
```{r}
COElkHunters <- summarise(group_by(COElkRifleAll,Year,Unit),
                          Hunters = sum(c(Hunters.Antlered,Hunters.Antlerless,Hunters.Either),na.rm = T))

COElkHunters$Year <- as.numeric(COElkHunters$Year)
```

Join in Hunter and Draw data together
```{r}
COElkHunters <- left_join(COElkHunters, COElkDraw, by = c("Year","Unit"))
```

Replace the draw data that don't have entries with 0
```{r}
COElkHunters$Drawn[is.na(COElkHunters$Drawn)] <- 0
COElkHunters$Quota[is.na(COElkHunters$Quota)] <- 0
```

Split into train and test sets. Will use 75% of the data to train on. Be sure to include
each unit in the split. ... so do the split for each unit, first make sure each Unit has
at least three entries

```{r}
COElkHunters <- mutate(group_by(COElkHunters,Unit),
                       numentries = n())
COElkHunters <- filter(COElkHunters, numentries >= 3)
COElkHunters$UnitYear <- paste(COElkHunters$Unit, COElkHunters$Year)

traindata <- COElkHunters %>% group_by(Unit) %>% sample_frac(size = .75, replace = F)
testdata <- COElkHunters[!COElkHunters$UnitYear %in% traindata$UnitYear,]

COElkHunters <- select(COElkHunters, -UnitYear, -numentries)

traindata <- select(traindata, -UnitYear, -numentries)
testdata <- select(testdata, -UnitYear, -numentries)
```


Save off for importing into AzureML
```{r}
write.csv(COElkHunters,file = "~/_code/colorado-dow/datasets/COElkHunters.csv",row.names = F)
```

notice that the number of hunters data is skewed.
```{r}
ggplot(COElkHunters, aes(Hunters)) + 
  geom_density() +
  xlab("Hunters in Unit") +
  ylab("Number of Units") +
  theme(axis.text.y = element_blank()) +
  labs(title="Distribution of Hunters in each Unit", subtitle="2006-2017", caption="source: cpw.state.co.us")

```


A general rule of thumb to consider is that skewed data whose ratio of the highest value to the 
lowest value is greater than 20 have significant skewness. Also, the skewness statistic can be 
used as a diagnostic. If the predictor distribution is roughly symmetric, the skewness values 
will be close to zero. As the distribution becomes more right skewed, the skewness statistic 
becomes larger. Similarly, as the distribution becomes more left skewed, the value becomes negative.
Replacing the data with the log, square root, or inverse may help to remove the skew.

Example of how BoxCox can redistribute the data
```{r}
preProcValues2 <- preProcess(as.data.frame(traindata), method = "BoxCox")
trainBC <- predict(preProcValues2, as.data.frame(traindata))
```

```{r}
ggplot(trainBC, aes(Hunters)) + 
  geom_density() +
  xlab("BoxCox Hunters in Unit") +
  ylab("Number of Units") +
  theme(axis.text.y = element_blank()) +
  labs(title="BoxCox Distribution of Hunters in each Unit", subtitle="2006-2017", caption="source: cpw.state.co.us")
```


```{r}
quickmethods <- c("lm",'svmLinear',"svmRadial","knn","cubist","kknn","glm.nb")

step1_all <- NULL
for (imethod in quickmethods) {
  step1 <- NULL
  start <- now()
  
  if (imethod == "lm") {
    controlmethod <- "repeatedcv"
  } else {controlmethod <- "adaptive_cv"}
  
  fitControl <- trainControl(
    method = controlmethod,
    # search = 'random',
    number = 4,
    repeats = 4,
    allowParallel = TRUE,
    summaryFunction = defaultSummary)
  
  registerDoSEQ()
  registerDoMC(cores = 6)
  
  HuntersModel_1 = train(Hunters ~ ., data = traindata,
                         method = imethod,
                         preProc = c("center","scale"), 
                         tuneLength = 15,
                         trControl = fitControl)
  
  HuntersModel_1
  
  # measure performance
  predictdata <- predict(HuntersModel_1, testdata)
  
  step1$method <- imethod
  step1$RMSE <- postResample(pred = predictdata, obs = testdata$Hunters)[1]
  step1$duration <- now() - start
  step1 <- as.data.frame(step1)
  step1_all <- rbind(step1_all,step1)
}
step1_all
```

